---
title: "Simple Linear Regression, Sum of Squares, and R Squared"
author: " Forest Kingfisher"
date: "`r Sys.Date()`"
output: html_document
---
## Exercise 2.12, taken fromMontgomery, Douglas C., et al. <u>Introduction to Linear Regression Analysis</u>, John Wiley & Sons, Incorporated, 2012.

The number of pounds of steam used per month at a plant is thought to be related to the average monthly ambient temperature. The past year’s usages and temperatures follow.

```{r, echo=TRUE}
#Import steam data by month
dat212 <- read.csv("https://raw.githubusercontent.com/forestwhite/Regression/main/data/212Table.csv")
names(dat212)[names(dat212) == "ï..Month"] <- "Month"
dat212
```

Generate a scatterplot of the data.

```{r}
#Generate a scatter plot of the data
plot(x = dat212$Temperature, 
     y = dat212$Usage.l000, 
     main = "Steam Usage vs. Ambient Temperature",
     xlab = "Ambient Temperature, Fahrenheit", 
     ylab = "Steam Usage, pounds", 
     xlim = c(20,80), 
     ylim = c(100,700))
```

Fit a simple linear regression model to the data and generate the intercept and coefficient estimates.  

```{r, echo=TRUE}
#generate linear model, scatter plot, and least squares coefficients
lm212<-lm(dat212$Usage.l000~dat212$Temperature)
summary(lm212)
```

Calculate the Sum of Squares Total (SST), the Sum of Squares of the Response (SSR), and the Sum of Squares of the Error/Residuals (SSE), and use them to find the coefficient of determination, R^2.

```{r}
# calculate y_hat estimate and sample average for y
yhat_i<-9.20847*dat212$Temperature-6.33209
y_bar<-mean(dat212$Usage.l000)

#Calculate the sum of squares values
SS_total<-sum((dat212$Usage.l000-y_bar)^2)
SS_total
SS_error<-sum((dat212$Usage.l000-yhat_i)^2)
SS_error
SS_regression<-sum((yhat_i-y_bar)^2)
SS_regression

# calculate coefficient of determination, R^2
r_squared<-SS_regression/SS_total
r_squared
```

### Complete Code

The complete R code used in this analysis is as follows:

```{r, eval=FALSE}
#Import steam data by month
dat212 <- read.csv("https://raw.githubusercontent.com/forestwhite/Regression/main/data/212Table.csv")
names(dat212)[names(dat212) == "ï..Month"] <- "Month"
dat212

#Generate a scatter plot of the data
plot(x = dat212$Temperature, 
     y = dat212$Usage.l000, 
     main = "Steam Usage vs. Ambient Temperature",
     xlab = "Ambient Temperature, Fahrenheit", 
     ylab = "Steam Usage, pounds", 
     xlim = c(20,80), 
     ylim = c(100,700))

#generate linear model, scatter plot, and least squares coefficients
lm212<-lm(dat212$Usage.l000~dat212$Temperature)
summary(lm212)

# calculate y_hat estimate and sample average for y
yhat_i<-9.20847*dat212$Temperature-6.33209
y_bar<-mean(dat212$Usage.l000)

#Calculate the sum of squares values
SS_total<-sum((dat212$Usage.l000-y_bar)^2)
SS_total
SS_error<-sum((dat212$Usage.l000-yhat_i)^2)
SS_error
SS_regression<-sum((yhat_i-y_bar)^2)
SS_regression

# calculate coefficient of determination, R^2
r_squared<-SS_regression/SS_total
r_squared
```