---
title: "Evaluate model adequecy, error normality, and variance of residuals"
author: " Forest Kingfisher"
date: "`r Sys.Date()`"
output: html_document
---
## Exercise 4.6, taken from Montgomery, Douglas C., et al. <u>Introduction to Linear Regression Analysis</u>, John Wiley & Sons, Incorporated, 2012.

The purity of oxygen produced by a fractional distillation process is thought to be related to the percentage of hydrocarbons in the main condenser of the processing unit. Twenty samples are shown below.

```{r, echo=TRUE}
#Import oxygen purity data
dat46 <- read.csv("https://raw.githubusercontent.com/forestwhite/Regression/main/data/46table.csv")
names(dat46)[names(dat46) == "Purity...."] <- "Purity"
names(dat46)[names(dat46) == "Hydrocarbon...."] <- "Hydrocarbon"
dat46
```

**a. Construct a normal probability plot of the residuals. Does there seem to be any problem with the normality assumption?**

Looking at the normal probability plot, the residuals indicate a moderately skewed/non-normal distribution, therefore there is evidence the data set may violate the assumption of normally distributed errors required for model adequacy.

```{r}
# load ggplot library
library(ggplot2)

# calculate least square estimate coefficients using linear model
lm46<-lm(dat46$Purity~dat46$Hydrocarbon)
coefficients(lm46)

# basic scatterplot with least squares estimate linear regression line
ggplot(data=dat46, aes(x=Hydrocarbon, y=Purity)) + 
  geom_point(color="blue") + 
  ggtitle("Purity (%) of oxygen samples vs.\n percentage hydrocarbon in main condenser")  + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  xlab("percentage of hydrocarbon") + 
  ylab("purity of oxygen (%)") +
  geom_abline(slope = coef(lm46)[["dat46$Hydrocarbon"]], 
              intercept = coef(lm46)[["(Intercept)"]],
              color="red")

# test normality of error terms: plot normal probability plot
plot(lm46, , which = 2)
```

**b. Construct and interpret a plot of the residuals versus the predicted response.**

To calculate the residuals and standardized residuals by hand:

 * Residuals/Error values are the difference between response values from the data and predicted response from linear model
 * Standardized residuals are the residuals/error values divided by the square root of the Mean Squared Error (MSE)

```{r}
# Calculate residuals/error values manually
Hydrocarbon<-dat46$Hydrocarbon
yhat_i<-11.80103*Hydrocarbon+77.86328
Errors <- dat46$Purity - yhat_i
dfErr<-data.frame(Hydrocarbon, Errors)
dfErr[order(dfErr$Hydrocarbon), ]

# Calculate standardized residuals manually
SS_error<-sum((dat46$Purity-yhat_i)^2)
MSE<-SS_error/(20-2) #approximate variance or sigma_squared of the sample
StandardizedResiduals<-Errors/(MSE^0.5)
dfStdRes<-data.frame(Hydrocarbon, StandardizedResiduals)
dfStdRes[order(dfErr$Hydrocarbon), ]
```

Looking at the residuals plotted against predicted values, they aren't distributed randomly, indicating that the data violates the assumption of constant variance (Homoscedasticity of errors) required for an adequate regression model. Standardized residuals show several outliers - observations 7:|-1.299|, 17:|1.45|, 18:|2.03| - which could be analyzed for leverage.

```{r}
# test constant variance: plot residuals vs. predicted values
plot(lm46, , which = 1)
# second test constant variance: plot standardized residuals vs. predicted values
plot(lm46, , which = 3)
```

### Complete Code

The complete R code used in this analysis is as follows:

```{r, eval=FALSE}
#Import oxygen purity data
dat46 <- read.csv("https://raw.githubusercontent.com/forestwhite/Regression/main/data/46table.csv")
names(dat46)[names(dat46) == "Purity...."] <- "Purity"
names(dat46)[names(dat46) == "Hydrocarbon...."] <- "Hydrocarbon"
dat46

# library
library(ggplot2)

# basic scatterplot with least squares estimate linear regression line
ggplot(data=dat46, aes(x=Hydrocarbon, y=Purity)) + 
  geom_point(color="blue") + 
  ggtitle("Purity (%) of oxygen samples vs.\n percentage hydrocarbon in main condenser")  + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  xlab("percentage of hydrocarbon") + 
  ylab("purity of oxygen (%)") +
  geom_abline(slope = coef(lm46)[["dat46$Hydrocarbon"]], 
              intercept = coef(lm46)[["(Intercept)"]],
              color="red")

# calculate least square estimate coefficients using linear model
lm46<-lm(dat46$Purity~dat46$Hydrocarbon)
coefficients(lm46)

# test normality of error terms: plot normal probability plot
plot(lm46, , which = 2)

# Calculate residuals/error values manually
Hydrocarbon<-dat46$Hydrocarbon
yhat_i<-11.80103*Hydrocarbon+77.86328
Errors <- dat46$Purity - yhat_i
dfErr<-data.frame(Hydrocarbon, Errors)
dfErr[order(dfErr$Hydrocarbon), ]

# Calculate standardized residuals manually
SS_error<-sum((dat46$Purity-yhat_i)^2)
MSE<-SS_error/(20-2) #approximate variance or sigma_squared of the sample
StandardizedResiduals<-Errors/(MSE^0.5)
dfStdRes<-data.frame(Hydrocarbon, StandardizedResiduals)
dfStdRes[order(dfErr$Hydrocarbon), ]

# test constant variance: plot residuals vs. predicted values
plot(lm46, , which = 1)
# second test constant variance: plot standardized residuals vs. predicted values
plot(lm46, , which = 3)
```